---
title: "DATA607_Week1"
author: "Nat Elkins"
date: "2/2/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

For this assignment, I chose FiveThirtyEight's MLB Elo dataset.  They use it in their [Complete History of MLB](https://projects.fivethirtyeight.com/complete-history-of-mlb/#) projcet as well as their [annual MLB predictions](https://projects.fivethirtyeight.com/2019-mlb-predictions/).  The Elo system is a comparative rating system that takes into account a team's wins and losses to determine their standing within a league (this system is used across many sports: golf, chess, basketball, etc.).  RStudio seems to crash when I try and load all 100+ years of baseball data, so let's just go with the latest season in 2019.

### Exploring the data

The MLB dataset can be found on FiveThirtyEight's website.  Let's load that data.

```{r load_mlb_elo_latest}
mlb_elo_latest <- read.csv("https://projects.fivethirtyeight.com/mlb-api/mlb_elo_latest.csv")
```

Let's take a look at what we've got.  First we can look at the dimensions of the data set.

```{r dimensions}
dim(mlb_elo_latest)
```

Length makes sense.  There's 30 teams in MLB and 162 games in the regular season.  (30 x 162) / 2 (since two teams participate in each game) = 2430 games, plus some postseason games leading up through to the World Series.  But what are all the columns?

```{r columns}
colnames(mlb_elo_latest)
```

Woah that's a lot.  Fortunately FiveThirtyEight gave us a nice description of each column:

Column | Definition
-----| ---------
date | Date of game
season | Year of season
neutral | Whether game was on a neutral site
playoff | Whether game was in playoffs, and the playoff round if so
team1 | Abbreviation for home team
team2 | Abbreviation for away team
elo1_pre | Home team's Elo rating before the game
elo2_pre | Away team's Elo rating before the game
elo_prob1 | Home team's probability of winning according to Elo ratings
elo_prob2 | Away team's probability of winning according to Elo ratings
elo1_post | Home team's Elo rating after the game
elo2_post | Away team's Elo rating after the game
rating1_pre | Home team's rating before the game
rating2_pre | Away team's rating before the game
pitcher1 | Name of home starting pitcher
pitcher2 | Name of away starting pitcher
pitcher1_rgs | Home starting pitcher's rolling game score before the game
pitcher2_rgs | Away starting pitcher's rolling game score before the game
pitcher1_adj | Home starting pitcher's adjustment to their team's rating
pitcher2_adj | Away starting pitcher's adjustment to their team's rating
rating_prob1 | Home team's probability of winning according to team ratings and starting pitchers
rating_prob2 | Away team's probability of winning according to team ratings and starting pitchers
rating1_post | Home team's rating after the game
rating2_post | Away team's rating after the game
score1 | Home team's score
score2 | Away team's score

I'm interested in seeing, at least in this season, what proportion of the time the Elo system accurately predicted the outcome of the game.  For that, we only need `elo_prob1`, or the probability that the first team would win (`elo_prob1` + `elo_prob2` = 1), and then the final scores of the game to determine who won.

```{r cut_down_columns}
columns_to_keep <- c("elo_prob1","score1","score2")
mlb_new <- mlb_elo_latest[columns_to_keep]
```

Let's also add two new columns to make life a bit easier.  We'll define one columns that represents whether the Elo score indicated team one would win, and then another that will represent whether they did win.

```{r add_new_columns}
mlb_new$team_one_will_win <- mlb_new$elo_prob1 > .5
mlb_new$team_one_won <- mlb_new$score1 > mlb_new$score2
mlb_new$prediction_accurate <- mlb_new$team_one_will_win == mlb_new$team_one_won
```

Then we can take the mean of the `prediction_accurate` column using the `mean()` function.  According to R documentation, `TRUE` and `FALSE` will be coerced to `1` and `0` respectively.

```{r take_the_mean}
mean(mlb_new$prediction_accurate)
```

## Conclusions

Hmm...59.20% doesn't seem _that_ much better than chance.  I'd be curious to examinee other predictive baseball models (of which there are many) and see if they fare any better.  Additionally, I'd like to see to what extent the accuracy of the Elo probability varies significantly from season to season, or even between teams.

